---
title: "Homework 5"
author: "Alex Soupir"
date: "October 13, 2019"
output:
  pdf_document:
    keep_md: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=F,warning=F,echo=F,fig_height=10,fig_width=7,cache = F)
```

*Packages*: vcd, lattice, randomForest, party, partykit, mboost, TH.data, ipred, rpart, randomForest, mlbench, tidyverse, boot, adabag, ggplot

*Collaborators*:

Please do the following problems from the text book R Handbook and stated.

<!--
1. Draw B bootstrap samples from the original data set i.e. we draw n out of n observattion with replacement from our n original observations
2. For the bootstrap sample grow a tree as follows: - At each node, randomly select q of the p predictors and restrict the splits based on the random subset of the q variables (q<<p)
3. Repeat the above 2 steps and generate a forest

Random forest: the above steps 1,2,3
Baggin is the above steps without step 2
Braggin is the same as baggin but take median instead of average
Bumping is the same as bagging but choose the best instead of average
Model aveageing is fit different models to original data and then take weighted averages based on same criterion (BIC/AIC)
Adaboost: using weights across trees rather than averages

cite seed

cp = complexity parameter


-->

1. The \textbf{BostonHousing} dataset reported by Harrison and Rubinfeld (1978) is available as data.frame package \textbf{mlbench} (Leisch and Dimitriadou, 2009). The goal here is to predict the median value of owner-occupied homes  (medv variable, in 1000s USD) based on other predictors in the dataset. Use this dataset to do the following 

    a.) Construct a regression tree using rpart(). The following need to be included in your discussion. How many nodes did your tree have? Did you prune the tree? Did it decrease the number of nodes? What is the prediction error (calculate MSE)?  Provide a plot of the predicted vs. observed values. Plot the final tree.
    
    ```{r regression_tree, echo=FALSE}
    library("vcd")
    library("lattice")
    library("randomForest")
    library("party")
    library("partykit")
    library("mboost")
    library("TH.data")
    library("ipred")
    library("rpart")
    library(mlbench)
    library(tidyverse)
    
    
    data(BostonHousing)
    bhous = BostonHousing
    
    #constructing trees
    set.seed(333)
    bhous_rpart = rpart( medv ~ ., data =bhous,
                         control=rpart.control(minsplit=10))
    
    opt <- which.min(bhous_rpart$cptable[,"xerror"])
    
    cp = bhous_rpart$cptable[opt, "CP"]
    bhous_prune = prune(bhous_rpart, cp = cp)
    #bhous_prune
    
    plot(as.party(bhous_rpart), 
         tp_args = list(id = FALSE))
    plot(as.party(bhous_prune),
         tp_args = list(id = FALSE))
    
    cat("Tree without Pruning: \n")
    medv_pred <- predict(bhous_rpart, newdata = bhous)
    single.mse = mean((bhous$medv - medv_pred)^2)
    single.mse
    cat("\nTree with Pruning: \n")
    medv_pred2 <- predict(bhous_prune, newdata = bhous)
    prune.mse = mean((bhous$medv - medv_pred2)^2)
    prune.mse
    
    xlim = range(bhous$medv)
    plot(medv_pred ~ medv, data = bhous, xlab = "Observed", 
         ylab = "Predicted", main="Unpruned Tree", ylim = xlim, xlim = xlim)
    abline(a = 0, b = 1)
    
    ggplot(bhous, aes(x=medv, y=medv_pred))+
      geom_point(shape=1)+
      geom_abline(slope=1, intercept=0)+
      labs(title="Unpruned Tree", x="Observed", y="Predicted")
    
    xlim = range(bhous$medv)
    plot(medv_pred2 ~ medv, data = bhous, xlab = "Observed", 
         ylab = "Predicted", main="Pruned Tree", ylim = xlim, xlim = xlim)
    abline(a = 0, b = 1)
    
    ggplot(bhous, aes(x=medv, y=medv_pred2))+
      geom_point(shape=1)+
      geom_abline(slope=1, intercept=0)+
      labs(title="Pruned Tree", x="Observed", y="Predicted")
    
    ```
    
    **I created a base tree as well as a pruned tree after setting seed to *333*. Both trees are using all parameters from the data frame to build the tree. Both trees had 8 internal nodes and 8 terminal nodes and both had the same MSE of 12.71556. Looking at the predicted vs observed values, the variance increases as the medv increases.**
    
    b) Perform bagging with 50 trees. Report the prediction error (MSE). Provide the predicted vs observed plot. 
    
    ```{r, echo=FALSE}
    set.seed(333)
    trees <- vector(mode = "list", length = 50)
    n <- nrow(bhous)
    bootsamples <- rmultinom(length(trees), n, rep(1, n)/n)
    mod <- rpart(medv ~ ., data = bhous, 
                 control = rpart.control(xval = 0))
    for (i in 1:length(trees))
      trees[[i]] <- update(mod, weights = bootsamples[,i])
    
    
    ## RP-glacoma - bagg
    #Closer look at the trees
    table(sapply(trees, function(x) 
      as.character(x$frame$var[1])))
    
    #Note the change in the choice of variables for the root node. 
    
    ## RP-glaucoma-splits
    classprob <- matrix(0, nrow = n, ncol = length(trees))
    for (i in 1:length(trees)) {
      classprob[,i] <- predict(trees[[i]], 
                               newdata = bhous)
      classprob[bootsamples[,i] > 0,i] <- NA
    }
    
    cat("\n50 Tree Bagging MSE: \n")
    avg <- rowMeans(classprob, na.rm = TRUE)
    fifty.mse = mean((bhous$medv - avg)^2)
    fifty.mse
    
    plot(avg ~ medv, data = bhous, xlab = "Observed", 
         ylab = "Predicted", ylim = xlim, xlim = xlim)
    abline(a = 0, b = 1)
    
    ggplot(bhous, aes(x=medv, y=avg))+
      geom_point(shape=1)+
      geom_abline(slope=1, intercept=0)+
      labs(title="50 Tree Ensemble", x="Observed", y="Predicted")
    ```
    
    **The MSE from bagging 50 trees is 17.02426 with seed *333*.**
    
    c) Use randomForest() function in R to perform bagging. Report the prediction error (MSE). Was it the same as (b)? If they are different what do you think caused it?  Provide a plot of the predicted vs. observed values.
    
    ```{r, echo=FALSE}
    library(randomForest)
    set.seed(333)
    rf = randomForest(medv ~ .,ntree=50, mtry = 13, data = bhous)
    rf
    rf.pred = predict(rf, newdata=bhous)
    
    cat("\nMSE from RandomForest() bagging with 50 trees: \n")
    rfb.mse = mean((bhous$medv - rf.pred)^2)
    rfb.mse
    
    plot(rf.pred ~ medv, data = bhous, xlab = "Observed", 
         ylab = "Predicted", ylim = xlim, xlim = xlim)
    abline(a = 0, b = 1)
    
    ggplot(bhous, aes(x=medv, y=rf.pred))+
      geom_point(shape=1)+
      geom_abline(slope=1, intercept=0)+
      labs(title="RF Bagging - 50 Tree", x="Observed", y="Predicted")
    ```
    
    **The random forest with *333* as the seed using all of the features (bagging by using *mtry = 13*) method provided a much lower MSE value than the 50 tree ensemble or the single tree (1.855 v 17.024 v 12.716). This is most likely due to the number of trees being created and used for predictions since the random forest function is creating 500 trees instead of 50 or just 1.**
    
    d) Use randomForest() function in R to perform random forest. Report the prediction error (MSE).  Provide a plot of the predicted vs. observed values.
    
    ```{r, echo=FALSE}
    library(randomForest)
    set.seed(333)
    rf2 = randomForest(medv ~ ., ntree=50, data = bhous)
    rf2
    rf2.pred = predict(rf2, newdata=bhous)
    
    cat("\nMSE from RandomForest() with 50 trees: \n")
    rf2.mse = mean((bhous$medv - rf2.pred)^2)
    rf2.mse
    
    plot(rf2.pred ~ medv, data = bhous, xlab = "Observed", 
         ylab = "Predicted", ylim = xlim, xlim = xlim)
    abline(a = 0, b = 1)
    
    ggplot(bhous, aes(x=medv, y=rf2.pred))+
      geom_point(shape=1)+
      geom_abline(slope=1, intercept=0)+
      labs(title="RF - 50 Tree", x="Observed", y="Predicted")
    ```
    
    **By creating a random forest with *333* as the seed and not telling it how many of the features to used to create the trees, the MSE went up slightly from bagging (2.047 v 1.855). Looking at the predicted vs observed plot, there were some values in the middle that moved closer to the x=y line, but values on the high end of medv spread out slightly more than with bagging, which increased the MSE.**
    
    e) Provide a table containing each method and associated MSE. Which method is more accurate?
    
    ```{r, echo=FALSE}
    tree.df = data.frame("Method" = c("Single Tree (un-pruned)",
                                      "Single Tree (pruned)",
                                      "50 Tree Ensemble",
                                      "Random Forest Bagging",
                                      "Random Forest"),
                         "Mean_Square_Error" = c(single.mse,
                                                 prune.mse,
                                                 fifty.mse,
                                                 rfb.mse,
                                                 rf2.mse))
    print(tree.df)
    ```
    
    **The most accurate method was random forest bagging with an MSE of 1.855. This is far lower than the 50 tree ensemble which has an MSE of 17.024. I suspect the 50 tree ensemble is performing worse due to a few poor trees.**
    
2. Consider the glacoma data (data = "\textbf{GlaucomaM}", package = "\textbf{TH.data}").

    a) Build a logistic regression model. Note that most of the predictor variables are highly correlated. Hence, a logistic regression model using the whole set of variables will not work here as it is sensitive to correlation.
        \begin{verbatim}
        glac_glm <- glm(Class ~., data = GlaucomaM, family = "binomial")
        #warning messages  -- variable selection needed 
        \end{verbatim}

        The solution is to select variables that seem to be important for predicting the response and using those in the modeling process using GLM. One way to do this is by looking at the relationship between the response variable and predictor variables using graphical or numerical summaries - this tends to be a tedious process. Secondly, we can use a formal variable selection approach. The $step()$ function will do this in R. Using the $step$ function, choose any direction for variable selection and fit logistic regression model. Discuss the model and error rate.
        
        \begin{verbatim}
        #use of step() function in R
        ?step
        glm.step <- step(glac_glm)
        \end{verbatim}
        
        
        Do not print out the summaries of every single model built using variable selection. That will end up being dozens of pages long and not worth reading through. Your discussion needs to include the direction you chose. You may only report on the final model, the summary of that model, and the error rate associated with that model.
    
    ```{r, echo=FALSE}
    set.seed(333)
    
    glau = GlaucomaM
    
    glau$Class = as.character(glau$Class)
    
    for(i in 1:length(glau$Class)){
        if(glau$Class[i] == "normal"){
            glau$Class[i] = 0
        } else {
            glau$Class[i] = 1
        }
    }
    
    glau$Class = as.factor(glau$Class)
    
    Class = glau$Class
    feat = glau[,-c(length(glau))]
    
    cors = cor(feat)
    cors[upper.tri(cors)] = 0
    diag(cors) = 0
    
    glau1 = feat[,!apply(cors,2,function(x) any(x > abs(0.5)))]
    glau1 = cbind(glau1, Class)
    
    glau.glm = glm(Class~., data=glau1, family="binomial")
    glm.step = step(glau.glm, trace=F)
    
    summary(glm.step)
    
    glau.pred = predict(glm.step, glau, type="response")
    glau.pred = ifelse(glau.pred >= 0.5, 1, 0)
    glau.obs = ifelse(glau$Class == "glaucoma", 1, 0)
    
    cat("\nMSE from step glm: \n")
    glmstep.mse = mean((glau.pred - glau.obs)^2)
    glmstep.mse
    
    ```
    
    **The direction that I chose to go given that there are many metrics that correlate with each other is to remove those with a correlation greater than 0.5. The remaining features were than used in the step() function to determine the best combination. The final model created used *phct*, *phci*, *hvc*, *mr*, *rnf*, and *emd*. All features were significant at P<=0.01 besides *phct*. The AIC of the model is 168.13 which is better than the other model created within the step() function with AIC=169.92. This indicates the final model fits the response variable (glaucoma or normal) better than the first model. The mean squared error of this final model is 0.485, which seems relatively low but without other models to compare to is hard to tell whether this is good or poor.**
    
    b) Build a logistic regression model with K-fold cross validation (k = 10). Report the error rate.
    
    ```{r, echo=FALSE}
    library(boot)
    set.seed(333)
    cost=function(r, pi=0)
      mean(abs(r-pi)>0.5)
    
    cat("K-fold cross validation (K=10) error rate: \n")
    glau.CV10 = cv.glm(glau, glau.glm, K=10, cost)$delta[1];glau.CV10
    ```
    
    **The error from cross validation is 0.27 which is almost half of the using step for glm. This indicates that k-fold cross validation better predicts whether the patient has glaucoma or not."**
    
    c) Find a function (package in R) that can conduct the "adaboost" ensemble modeling. Use it to predict glaucoma and report error rate. Be sure to mention the package you used.
    
    ```{r, echo=FALSE}
    library(adabag)
    set.seed(333)
    ind = sample(2, nrow(glau), replace = TRUE, prob = c(0.8, 0.2))
    train.df = glau[ind == 1,]
    test.df = glau[ind == 2,]
    
    glau.boost = boosting(glau.glm, train.df, control=rpart.control(minsplit=10))
    glau.b.pred = predict.boosting(glau.boost, newdata=test.df)
    boost.preds = as.numeric(glau.b.pred$class)
    glau.test.obs = ifelse(test.df$Class == "1", 1, 0)
    
    glau.boostc = boosting(glau.glm, glau)
    glau.b.predc = predict.boosting(glau.boostc, newdata=glau)
    boost.predsc = as.numeric(glau.b.predc$class)
    glau.test.obsc = ifelse(glau$Class == "1", 1, 0)
    
    cat("MSE from whole data boosting: \n")
    boost.msec = mean((boost.predsc - glau.test.obsc)^2)
    boost.msec
    
    cat("\nMSE from train boosting: \n")
    boost.mse = mean((boost.preds - glau.test.obs)^2)
    boost.mse
    ```
    
    **Created a 80:20 train:test split because the model was perfectly fitting the data. With predicting if the patient has glaucoma the boosting model has an MSE of 0.195. This is better than both step glm and k-fold cross validation. I suspect this is due to the 100 trees that it uses to build the trees and it calculating the weight of each tree based on how well it performs. This is with using a control of *minsplit=10* as per the documentation for the *boosting* funciton.**
    
    d) Report the error rates based on single tree, bagging and random forest. (A table would be great for this).
    
    ```{r, echo=FALSE}
    #single tree
    set.seed(333)
    glau1$Class = as.character(glau1$Class)
    glau1$Class = as.numeric(glau1$Class)
    
    glau1_rpart = rpart(Class ~ ., data =glau1,
                         control=rpart.control(minsplit=10))
    
    opt <- which.min(glau1_rpart$cptable[,"xerror"])
    
    cp = glau1_rpart$cptable[opt, "CP"]
    glau1_prune = prune(glau1_rpart, cp = cp)
    
    class_pred <- predict(glau1_prune, newdata = glau1)
    
    single1.mse = mean((glau1$Class - class_pred)^2)
    
    xlim = range(glau1$Class)
    
    #bagging
    set.seed(333)
    class.trees <- vector(mode = "list", length = 50)
    n <- nrow(glau1)
    bootsamples <- rmultinom(length(class.trees), n, rep(1, n)/n)
    mod <- rpart(Class ~ ., data = glau1, 
                 control = rpart.control(minsplit=10))
    for (i in 1:length(class.trees))
      class.trees[[i]] <- update(mod, weights = bootsamples[,i])
    
    classprob <- matrix(0, nrow = n, ncol = length(class.trees))
    for (i in 1:length(class.trees)) {
      classprob[,i] <- predict(class.trees[[i]], 
                               newdata = glau1)
      classprob[bootsamples[,i] > 0,i] <- NA
    }
    
    avg <- rowMeans(classprob, na.rm = TRUE)
    fifty.mse2 = mean((glau1$Class - avg)^2)
    
    #RF bagging
    set.seed(333)
    rfc = randomForest(Class ~ ., mtry = 8, ntree=50, data = glau1)
    
    rfc.pred = predict(rfc, newdata=glau1)
    
    rfc.mse = mean((glau1$Class - rfc.pred)^2)
    
    
    #RF
    set.seed(333)
    rfc2 = randomForest(Class ~ ., ntree=50, data = glau1)
    
    rfc.pred2 = predict(rfc2, newdata=glau1)
    
    rfc.mse2 = mean((glau1$Class - rfc.pred2)^2)
    
    
    
    
    
    
    model.df = data.frame("Method" = c("Single Pruned Tree",
                                      "Bagging-50",
                                      "RF Bagging-50",
                                      "RF-50"),
                         "Mean_Square_Error" = c(single1.mse,
                                                 fifty.mse2,
                                                 rfc.mse,
                                                 rfc.mse2))
    print(model.df)
    ```
    
    e) Write a conclusion comparing the above results (use a table to report models and corresponding error rates). Which one is the best model?
    
    **Of the results for a single tree, bagging with 50 trees, random forest bagging with 50 trees, and random forest with 50 trees, the random forest bagging has the lowest error rate with a mean squared error of 0.028. The bagging with 50 trees had the highest MSE with 0.147, which is higher than even the single pruned tree which had an MSE of 0.117. Not telling the random forest to pick from all 8 features in the non-correlated dataframe performed slightly worse according to the MSE than did the *randomForest()* bagging method.**

    f) From the above analysis, which variables seem to be important in predicting Glaucoma?
    
    ```{r, echo=FALSE}
    message("Boosting Feature Importance:")
    glau.boost$importance
    ```
    
    **From the above analysis, the features that seem to be important are those included in creating the models; *phct*, *phci*, *hvc*, *mr*, *rnf*, and *emd*. The glm model gave the significance of the coefficients for each and *phct* wasn't significantly different from zero. The boosting function provides the importance of the features used in making the trees and *rnf* was the most important with *mv* being the least important.**

*Resources Used*:

+ StackOverflow
+ datascienceplus.com
+ rdocumentation.org
